Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-30 doc/paper-lcr.tex
working on intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Pangloss: Testing the reconstruction paper
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title[Line of Sight Mass Reconstruction]
{Inferring Lensing Mass in the Universe \\
from Photometric Catalog Data}

@@

@@

\begin{abstract} 
Strong gravitational lenses can provide high precision cosmological distance
measurements; one of the most important sources of systematic error in these
estimates is the additional mass along the line of sight to the source. Weak
lensing perturbations by galaxies and groups in this beam can contribute an ``external
convergence'' $\kappa$ that dominates the  uncertainty in the inferred
distances; likewise, it may also contaminate estimates of the luminosity of
objects at high redshift (such as reionisation-era proto-galaxies, and type Ia
supernovae standard candles).  We characterise this uncertainty by marginalising
over a probability density function for $\kappa$: here, we investigate the use
of a simple halo model to estimate $\Pr(\kappa)$ given noisy estimates of the
photometric redshift and stellar mass of galaxies observed in optical imaging
surveys of the field in question. We use mock catalogs from the Millenium
Simulation and compare our reconstructed $\Pr(\kappa)$ to ray-traced $\kappa$ values,
we find that our model produces a precise estimator of $\kappa$ with only small biases.
For each lightcone the individual $\Pr(\kappa)$ from this reconstruction process are
typically 1.8 times more precise than the global distribution of $\kappa$ given perfect knowledge 
of the halo mass and redshift. After incorporating uncertainties on redshift and halo and stellar masses
we typically find that the reconstruction yields an improvement factor of 1.4. For each line of sight the
 $\Pr(\kappa)$ can be formed before the investment of light-curve monitoring: by following up only the 
most costrained lines of sight a factor of 2 improvement over the global $\pr(\kappax)$ should be 
achievable.
\end{abstract}

@@

Every distant object we observe has had its shape distorted, and size
and total brightness magnified (or demagnified) by a compound lens constructed
from all the mass distributed between us and it. As \citet{Vale+White2003} and \citet{HilbertEtal2007}
showed, there are no empty lightcones in our universe. This fact makes gravitational lensing a potentially important source of systematic error for any estimate of luminosity or distance, an issue
raised for \eg type Ia supernovae by \citet[][]{Holz+Wald1998,Linder+Holz2004}, for
gamma ray bursts by \citet[][]{Oguri+Takahashi2006,Wang+Dai2011}, 
and for high redshift galaxies by \citet{BradacEtal2009}, among others.

Along lines of sight containing strong gravitational lenses,
the effect has been found to be
particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios \citep[][]{DalalEtal2005}.
\citet{SuyuEtal2010} found that in time delay lens cosmography the
lensing effect of mass structures along the (unusually over-dense) line of
sight to the quadruply-imaged radio source B1608$+$656, the so-called ``external
convergence'' $\kappax$, was the dominant source of uncertainty in 
their 6\% measurement of Hubble's constant.

The problem is that one simply doesn't know
how magnified and distorted the source has been because the unlensed
source is unobservable. The external shear can potentially be recovered
in the mass-modelling of the lens, but the convergence is undetectable
from the image positions, shapes and relative fluxes; this is the
famous {\emph{  mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

How can we correct for this perturbative lensing and make
accurate measurements through the Universe? When inferring global quantities
(such as the cosmological parameters), averaging the results from a large
number of independent objects will tend to reduce the uncertainty, as the
ensemble-averaged convergence approaches zero for an unbiased sample
\citep[\eg][]{Linder+Holz2004}. If the sample is biased, however, some
residual systematic error will remain while the statistical uncertainty
decreases -- and this will be the case for any sample selected by the
brightness of the lensed object, or any other quantity that depends on
$\kappa$. Large
samples of lenses are expected to be discovered in the next decade in
ground-based optical imaging surveys \citep{Oguri+Marshall2010} -- but these
samples will be image separation-limited, due to the surveys' relatively poor
($\sim 1$~arcsec) image quality. There is a possibility that
they are somewhat biased towards high values of $\kappa$, although 
\citet{Fassnacht+Lubin2002} \todo{Anyone}{check this is the right Fassnacht paper} found no evidence of lenses lying along over-dense lines of sight.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Pangloss: Testing the reconstruction method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title[Line of Sight Mass Reconstruction]
{Reconstructing the Lensing Mass in the Universe \\
from Photometric Catalog Data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}

\todo{PJM}{Make abstract match conclusions.}

Strong gravitational lenses can provide high precision cosmological distance
measurements; one of the most important sources of systematic error in these
estimates is the additional mass along the line of sight to the source. Weak
lensing perturbations by galaxies and groups in this beam can contribute an ``external
convergence'' $\kappa$ that dominates the  uncertainty in the inferred
distances; likewise, it may also contaminate estimates of the luminosity of
objects at high redshift (such as reionisation-era proto-galaxies, and type Ia
supernovae standard candles).  We characterise this uncertainty by marginalising
over a probability density function for $\kappa$: here, we investigate the use
of a simple halo model to estimate $\Pr(\kappa)$ given noisy estimates of the
photometric redshift and stellar mass of galaxies observed in optical imaging
surveys of the field in question. We use mock catalogs from the Millenium
Simulation and compare our reconstructed $\Pr(\kappa)$ to ray-traced $\kappa$ values,
we find that our model produces a precise estimator of $\kappa$ with only small biases.
For each lightcone the individual $\Pr(\kappa)$ from this reconstruction process are
typically 1.8 times more precise than the global distribution of $\kappa$ given perfect knowledge 
of the halo mass and redshift. After incorporating uncertainties on redshift and halo and stellar masses
we typically find that the reconstruction yields an improvement factor of 1.4. For each line of sight the
 $\Pr(\kappa)$ can be formed before the investment of light-curve monitoring: by following up only the 
most constrained lines of sight a factor of 2 improvement over the global $\pr(\kappax)$ should be 
achievable.

\end{abstract}

@@

Every distant object we observe has had its apparent shape distorted,
and size and total brightness magnified (or demagnified) by a compound
weak gravitational lens constructed from all the mass distributed
between us and it. As \citet{Vale+White2003} and \citet{HilbertEtal2007}
showed, there are no empty lines of sight through our universe. This
fact makes gravitational lensing a potentially important source of
systematic error for any estimate of luminosity (or distance), an issue
raised for \eg type Ia supernovae by
\citet[][]{Holz+Wald1998,Linder+Holz2004}, for gamma ray bursts by
\citet[][]{Oguri+Takahashi2006,Wang+Dai2011},  and for high redshift
galaxies by \citet{BradacEtal2009}, among others.

Along lines of sight containing strong gravitational lenses, the
perturbative effects of line of sight mass structure have been found to
be particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios
\citep[][]{DalalEtal2005}. Indeed, \citet{SuyuEtal2010} found that in
time delay lens cosmography the so-called ``external convergence''
$\kappa$ due to mass structures along the (unusually over-dense) line
of sight to the quadruply-imaged radio source B1608$+$656 was the
dominant source of uncertainty in their 6\% measurement of Hubble's
constant. The problem is that one simply doesn't know how magnified and
distorted the source has been, because the unlensed source is
unobservable. The external shear can potentially be recovered in the
mass-modelling of the lens, but the convergence is undetectable from the
image positions, shapes and relative fluxes; this is the famous {\emph{ 
mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

How can we correct for this perturbative lensing and make accurate
measurements through the Universe? When inferring global quantities
(such as the cosmological parameters), combining the results from a
large number of independent objects will tend to reduce the uncertainty
due to external convergence, as the ensemble-averaged magnification
approaches zero for an unbiased sample \citep[\eg][]{Holz+Linder2005}.
If the sample is biased, however, some residual systematic error will
remain even as the statistical uncertainty decreases. Lensing bias will
be present (to varying degrees) in any sample {\it selected} by
brightness (which includes magnification), or any other quantity that
depends on $\kappa$. This bias may be introduced either at the object
detection stage, or later on when making decisions about which objects
to study further.  Moreover, we expect these biases to be strongest in
the strong lens systems, where magnification is highest. Large samples
of lenses are expected to be discovered in the next decade in
ground-based optical imaging surveys \citep{Oguri+Marshall2010}: both
the detectability of these lenses,  and the selection of sub-samples of
them for further observation, could be sensitive to the external
convergence.

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected by measuring the
small distortions it induces on the images of many background galaxies
in the field. \todo{PJM}{Write about Bernstein project in 0957.}

\todo{PJM}{Rearrange following section. After lunch.}
